Assessment of a Shen-Fourier turbulent channel flow solver for large-scale simulations

### 2016-03-17 paper.tex
Work on getting all algorithms in there

Three-dimensional basis functions and function spaces that are periodic in $y$ 
and $z$ directions, can now be defined as
\begin{align}
  \psi_{\bm{k}}(\bm{x}) = \phi_{l}(x)e^{ \imath(\underline{m} y + \underline{n} z)}, \quad V_N &= \text{span} \{ \psi_{\bm{k}}(\bm{x}):\, \bm{k} \in K_N  \}, \\
  \D{\psi}_{\bm{k}}(\bm{x}) = \D{\phi}_{l}(x)e^{ \imath(\underline{m} y + \underline{n} z)}, \quad \D{V}_N &= \text{span} \{ \D{\psi}_{\bm{k}}(\bm{x}):\, \bm{k} \in \D{K}_N  \}, \\
  \N{\psi}_{\bm{k}}(\bm{x}) = \N{\phi}_{l}(x)e^{ \imath(\underline{m} y + \underline{n} z)}, \quad \N{V}_N &= \text{span} \{ \N{\psi}_{\bm{k}}(\bm{x}):\, \bm{k} \in \N{K}_N  \},
\end{align}
where $\imath=\sqrt{-1}$ and 
\begin{align}
K_N = \Big\{\bm{k} \in \mathcal{Z}\times\mathcal{R}^2 / &(l, \underline{m}, \underline{n}) \in \left( l, \frac{2 \pi m}{L_y}, \frac{2 \pi n}{L_z} \right), \text{where} \notag \\
 &(l, m, n) \in  [0, \ldots, N_x] \times [-\frac{N_y}{2},\ldots, \frac{N_y}{2}-1] \times[-\frac{N_z}{2},\ldots, \frac{N_z}{2}-1] \Big\}. \label{eq:wavenumbermesh}
\end{align}
$\D{K}_N$ and $\N{K}_N$ are both equal to $K_N$ except from index $l$ that 
ranges $[0, 1, \ldots, N_x-2]$ and $[0, 1, \ldots, N_x-4]$ for $\D{K}_N$ 
and $\N{K}_N$ respectively, see (\ref{eq:phiD}) and (\ref{eq:phiN}). We use 
notation $K_N^p(m, n)=K_N(0, m, n)$ when referring to only the two periodic 
dimensions of $K_N$.

The computational mesh has in real space been created using $N = (N_x, N_y, 
N_z)$ 
intervals, where the two periodic directions use uniform intervals. The 
computational mesh is given as
\begin{align}
  X_N = \Big\{ \bm{x} \in \mathcal{R}^3/&(x_i, y_j, z_k) \in \left( h(i), \frac{jL_y}{N_y}, \frac{kL_z}{N_z} \right), \text{where} \notag \\
  &(i, j, k) \in [0, 1, \ldots, N_x] \times [0, 1, \ldots, N_y-1] \times [0, 1, \ldots, N_z-1] \Big\} \label{eq:Xn}
\end{align}
where $x_i = h(i)$ represents either
\begin{equation}
 h(i) = \begin{cases}
   \cos \left(\frac{i \pi }{N_x} \right) \, &\forall \, i=0,1, \ldots, N_x \quad  \text{for Gauss-Lobatto points}, \\
   \cos \left(\frac{(2i +1)\pi}{2N_x+2} \right) \, &\forall \, i=0,1, \ldots, N_x \quad  \text{for Gauss-Chebyshev points}. \\
 \end{cases}
\end{equation}

@@

If the expansion coefficients $\hat{u}$  are known for the entire wavenumber 
mesh (\ref{eq:wavenumbermesh}), then the expression (\ref{eq:u_solx}) may be 
evaluated on the mesh (\ref{eq:Xn}) using fast Fourier and fast Shen inverse 
transforms as
\begin{align}
u(x_i, y_j, z_k, t) &= \underbrace{\frac{1}{N_z}\sum_n 
\underbrace{\frac{1}{N_y} \sum_m \underbrace{\sum_l 
\hat{u}(l,m,n)\N{\phi}_l(x_i)}_{\N{\mathcal{S}}_x^{-1}} e^{\imath \underline{m} 
y_j}}_{\mathcal{F}_y^{-1}} e^{\imath \underline{n} z_k}}_{\mathcal{F}_z^{-1}} , 
\quad \forall \, \bm{x} \in X_N, \notag \\
  u &= \N{\mathcal{T}}^{-1}(\hat{u}) =  \mathcal{F}^{-1}_z (\mathcal{F}^{-1}_y 
  (\N{\mathcal{S}}^{-1}_x (\hat{u}))), \quad \forall \, \bm{x} \in X_N.  
  \label{eq:ifft} 
\end{align}
Here $\N{\mathcal{T}}^{-1}$ is used as short notation for the complete inverse 
transform, $\mathcal{F}_{y}^{-1}$ and $\mathcal{F}_{z}^{-1}$ represents inverse 
Fourier transforms along directions $y$ and $z$ respectively, and the inverse 
Shen transform $\N{\mathcal{S}}_{x}^{-1}$ is performed along the wall normal 
$x$ direction. Note that the transforms are applied sequentially to the entire 
computational mesh. The inverse Shen transform may be computed using fast 
Chebyshev transforms for all the three bases in (\ref{eq:Tk}, \ref{eq:phiD}, 
\ref{eq:phiN}). The transforms are slightly different for the three bases and 
${\mathcal{S}}, \D{\mathcal{S}}$ and $\N{\mathcal{S}}$ are used to distinguish 
between them with obvious notation.

@@

The scalar product  $h\N{\mathcal{S}}$ in (\ref{eq:sst1}) does not represent a 
complete transform. To find a transformation between physical $u$ and spectral 
$\hat{u}$ we make use of Eq. (\ref{eq:u_solx}) directly and use the discrete 
orthogonality of the Fourier basis functions
\begin{align}
\left<u, \N{\psi}_{\bm{k}}\right>_w &= \frac{1}{N_yN_z}\left< 
\sum_{\bm{m}=(q,r,s) \in \N{K}_N} \hat{u}_{\bm{m}} \N{\psi}_{\bm{m}}, 
\N{\psi}_{\bm{k}} \right>_w, \notag \\
           &= h \sum_{q=0}^{N_x-4} \sum_{i=0}^{N_x} \N{\phi}_{q}(x_i) 
           \N{\phi}_{l}(x_i) w_i \, \hat{u}(q, {m}, {n}, t), \notag \\
           &= h \sum_{p=0}^{N_x-4} \N{B}_{lq} \hat{u}(q, {m}, {n}, t), 
           \label{eq:sst2}
\end{align}
where $\N{B}_{lq} = (\N{\phi}_q, \N{\phi}_l)_w = 
\sum_{i=0}^{N_x} \N{\phi}_{q}(x_i) \N{\phi}_{l}(x_i) w_i$ is a 
pentadiagonal mass matrix. The complete transformation is now obtained by 
setting Eq. (\ref{eq:sst2}) equal to (\ref{eq:sst1}) and solving for $\hat{u}$. 
Moving to matrix notation we get
\begin{align}
h\sum_{q=0}^{N_x-4} \N{B}_{lq}\, \hat{u}(q, {m}, {n}, t) &= 
h\N{\mathcal{S}}(u)(l, m, n, t), \notag \\
 \hat{u}_{\bm{k}}(t) &= \N{\mathcal{T}}(u) =  \N{B}^{-1}\N{\mathcal{S}}(u), 
 \quad \forall \, \bm{k} \in \N{K}_N. \label{eq:fstB}
\end{align}
where $\N{\mathcal{T}}(u)$ denotes the complete transformation, such that $u = 
\N{\mathcal{T}}^{-1}(\N{\mathcal{T}}(u))$. Note that since $\N{B}$ assembles to 
a pentadiagonal matrix the solution ($\N{B}^{-1}$) can be obtained very fast 
and 
the complete 
transformation thus requires a fast Chebyshev transform ($O(N_x \log N_x)$) and 
a fast linear algebraic solve ($O(N_x)$). Similar transforms $\mathcal{T}$ 
and $\D{\mathcal{T}}$ are defined for the two other bases (\ref{eq:Tk}) and 
(\ref{eq:phiD}), using mass matrices $B_{lq}=(T_q, T_l)_w$ (diagonal) and 
$\D{B}_{lq}=(\D{\phi}_l, \D{\phi}_q)_w$ (tridiagonal) and scalar products 
$\mathcal{S}_l(\cdot) = (\cdot, T_l)_w$ and $\D{\mathcal{S}}_l(\cdot) = (\cdot, 
\D{\phi}_l)_w$.

@@

It remains to discretize the equations in time. Kim, Moin and Mansour use 
Crank-Nicolson for the diffusive terms and Adams-Bashforth for the nonlinear 
convective terms, whereas a third order Runge-Kutta method is used for the 
nonlinear term by Moser, Kim and Mansour (1999). We have implemented both 
approaches, but for simplicity only the first one is described here. 
Discretization in time is performed using a constant time step $\triangle t$, 
such that the time is represented discretely as $t_n = n \triangle t, \, n \in 
\mathcal{Z}^{\ge}$. After 
discretizing in time, equations (\ref{eq:u1}) and (\ref{eq:g1}) are transformed 
to a matrix form that can be solved for $\hat{u}^{n+1}$ and $\hat{g}^{n+1}$ 
given wavenumbers $m$ and $n$
\begin{align}
\N{H}\hat{u}^{n+1} & = \left(2\N{A} - 2\underline{z}^2\N{B} - \N{H} 
\right)\hat{u}^{n} + \N{\mathcal{S}}(h_v^{n+1/2}) \triangle t & 
\forall \, {\bm{k}} \in \N{K}_N, 
\label{eq:ufin}\\ 
\D{H} \hat{g}^{n+1} &= \left(2 \D{B}-\D{H}\right) 
\hat{g}^{n} + \D{\mathcal{S}}(h_g^{n+1/2}) \triangle t &\forall \, 
{\bm{k}} \in 
\D{K}_N. \label{eq:gfin}
\end{align}
The convective terms are computed with 
a forward Adams-Bashforth projection, such that $\N{\mathcal{S}}(h_v^{n+1/2}) = 
\N{\mathcal{S}}(1.5h_v^{n} - 0.5 h_v^{n-1})$ and $\D{\mathcal{S}}(h_g^{n+1/2}) 
= \D{\mathcal{S}}(1.5h_g^{n} - 0.5 h_g^{n-1})$ and the coefficient matrices are 
given as
\begin{align}
\N{H}_{lq}(m, n) &= -\frac{\nu \triangle t}{2}\N{Q}_{lq} + \left( 1 + \nu 
\triangle t 
\underline{z}^2 \right) \N{A}_{lq} - \frac{2\underline{z}^2 + \nu \triangle t 
\underline{z}^4}{2} \N{B}_{lq} &\forall m, n \in K_N^p, \\
\D{H}_{lq}(m, n) &= -\frac{\nu \triangle 
t}{2}\D{A}_{lq} + (1 + \frac{\nu \triangle t \underline{z}^2}{2}) \D{B}_{lq} 
&\forall m, n \in K_N^p,
\end{align}
where $\underline{z}^2 = \underline{m}^2 + \underline{n}^2$. Equations 
(\ref{eq:v0}) and (\ref{eq:w0}) are also discretized in time using 
Crank-Nicolson and Adams-Bashforth, and we obtain
\begin{align}
\left(-\frac{\nu \triangle t}{2}\D{A} + \D{B} \right)\hat{v}_0^{n+1} &= 
\left(\frac{\nu \triangle t}{2}\D{A} + \D{B} 
\right)\hat{v}^{n}_0 + \triangle t\D{B}\hat{\mathcal{H}}_{y,0}^{n+1/2} , 
\label{eq:v00} \\
\left(-\frac{\nu \triangle t}{2}\D{A} + \D{B} \right)\hat{w}_0^{n+1} &= 
\left(\frac{\nu \triangle t}{2}\D{A} + \D{B} 
\right)\hat{w}^{n}_0 + \triangle t\D{B}\hat{\mathcal{H}}_{z,0}^{n+1/2}, 
\label{eq:w00}
\end{align}
where we have used notation $\hat{v}^{n}_0 = \hat{v}(l, 0, 0, t_n)$, 
$\hat{\mathcal{H}}^{n}_{y,0} = \hat{\mathcal{H}}_y(l, 0, 0, t_n)$ and similar 
for $\hat{w}^n_0$ and $\hat{\mathcal{H}}^n_{z,0}$.

@@

The biharmonic operator $\N{H}$ is more challenging to solve efficiently, but 
it is still possible to find an algorithm that is $O(N)$ for a system of $N$ 
unknowns. Note 
that $\N{H}$ is the sum of three matrices $\N{Q}, \N{A}$ and $\N{B}$ and can be 
written as
\begin{equation}
\N{H} = c_0\N{Q} + c_1\N{A} + c_2 \N{B},
\end{equation}
where $c_0, c_1$ and $c_2$ are constants. The matrix $\N{A}_{kj}$ contains only 
three nonzero diagonals at $ j = k-2, k, k+2$,  whereas matrix $\N{B}_{kj}$ 
contains five 
nonzero diagonals at $ j = k-4, k-2, k, k+2, k+4$. The nonzero elements of the 
$\N{Q}_{kj}$ matrix are given as
\begin{align}
 \N{Q}_{kk} &= -4(k+1)(k+2)^2 \\
 \N{Q}_{kj} &= p_kq_j + r_ks_j, \quad j = k+2, k+4, \ldots N_x-4
\end{align}
where 
\begin{align}
p_k &= 8 k (k+1)(k+2)(k+4)\pi, &q_j = \frac{1}{j+3} \label{eq:pk} \\
r_k &= 24(k+1)(k+2)\pi, &s_j =  \frac{(j+2)^2}{(j+3)} \label{eq:rk}
\end{align} 
For each row in $\N{H}_{kj}$ we have that $\N{H}_{kj}=c_0\N{Q}_{kj}\, \forall j 
> k+4$. A direct LU decomposition of $\N{H}$ can be performed as shown in Alg. 
\ref{alg:lu_biharmonic}, which leads to a lower triangular 
matrix $\N{L}_{kj}$ with two nonzero diagonals at $j=k-2$ and $k-4$ plus a 
unity 
main diagonal. The upper triangular matrix $\N{U}_{kj}$ contains three distinct 
diagonals at $j=k, k+2$ and $k+4$. The remaining part of $\N{U}_{kj}$ can be 
written as
\begin{equation}
\N{U}_{kj} = a_k q_j + b_k s_j, \quad j = k+6, k+8, \ldots, N_x-4 \label{eq:ab}
\end{equation}
where $\bm{a}$ and $\bm{b}$ are two new vectors that can be computed 
recursively as shown in Alg. (\ref{alg:ab}). If $\bm{a}$ and $\bm{b}$ are 
pre-computed 
the total storage requirement for the complete LU decomposition is no more than 
$7N$ and the solution of the complete system can be obtained very quickly with 
one forward elimination and a back solve. The back 
solve is very fast because of the recursive formulation (\ref{eq:ab}), leading 
to a formula like (valid for $k \le N_x-8$)
\begin{equation}
u_k = \left(y_k - \N{U}_{k, k+2} u_{k+2} - \N{U}_{k, k+4} u_{k+4} - 
a_k\sum_{\substack{j=k+6\\k-j \text{ even}}}^{N_x-4} q_j u_j - 
b_k\sum_{\substack{j=k+6\\k-j \text{ even}}}^{N_x-4} 
s_j u_j\right)/\N{U}_{kk},
\end{equation}
that only requires 
one new addition to the row-sums for each back-traversed
row, see Alg. (\ref{alg:SolveBiharmonic}).

\begin{algorithm}
	\caption{LU decomposition of biharmonic operator $\N{H}$. Current algorithm 
	is using dense storage of $U_{kj}$, but no more than the 3 diagonals need 
	to be stored.}
	\label{alg:lu_biharmonic}
	\begin{algorithmic}[1]
		\Procedure{LUbiharmonic}{}
			\State \textbf{Input:} $\{\N{H}_{kj}\}_{k, j = 0}^{N}$
	        \State $U_{0j} = \N{H}_{0j} \quad \forall \, j \in 0, 2, \ldots, N$
	        \State $U_{1j} = \N{H}_{1j} \quad \forall \, j \in 1, 3, \ldots, N$
	        \State $L_{20} = \N{H}_{20}/U_{00}$
	        \State $L_{31} = \N{H}_{31}/U_{11}$
	        \State $U_{2j} = L_{20} U_{0j} \quad \forall \, j \in 2, 4, \ldots, 
	        N$
	        \State $U_{3j} = L_{31} U_{1j} \quad \forall \, j \in 3, 5, \ldots, 
	        N$
	        \For{$k = 4, 5, \ldots, N$}
	            \State $L_{k,k-2} = {U}_{k,k-2}/{U}_{k-2, k-2}$
		        \State $U_{kj} = {U}_{kj} - L_{k, k-2} U_{k-2, j} \quad 
		        \forall j = k, k+2, \ldots, N$
		        \If{$k <= N-2$}
		        \State $L_{k+2,k-2} = \N{H}_{k+2, k-2} / U_{k-2, k-2}$
		        \State $U_{k+2,j} = \N{H}_{k+2,j} - L_{k+2, k-2} U_{k-2, j} 
		        \quad \forall j = k-2, k, \ldots, N$
		        \EndIf
	        \EndFor
	        \State \Return $\{L_{kj}\}_{k, j = 0}^{N}$, $\{U_{kj}\}_{k, j = 
	        0}^{N}$
		\EndProcedure	
	\end{algorithmic}
\end{algorithm}

>>>>

Three-dimensional basis functions and function spaces that are periodic in $y$ 
and $z$ directions, can now be defined as
\begin{align}
  \psi_{\bm{k}}(\bm{x}) = \phi_{l}(x)e^{ \imath(\underline{m} y + \underline{n} z)}, \quad V_N &= \text{span} \{ \psi_{\bm{k}}(\bm{x}):\, \bm{k} \in K_N  \}, \\
  \D{\psi}_{\bm{k}}(\bm{x}) = \D{\phi}_{l}(x)e^{ \imath(\underline{m} y + \underline{n} z)}, \quad \D{V}_N &= \text{span} \{ \D{\psi}_{\bm{k}}(\bm{x}):\, \bm{k} \in \D{K}_N  \}, \\
  \N{\psi}_{\bm{k}}(\bm{x}) = \N{\phi}_{l}(x)e^{ \imath(\underline{m} y + \underline{n} z)}, \quad \N{V}_N &= \text{span} \{ \N{\psi}_{\bm{k}}(\bm{x}):\, \bm{k} \in \N{K}_N  \},
\end{align}
where $\imath=\sqrt{-1}$ and 
\begin{align}
K_N = \Big\{\bm{k} \in \mathcal{Z}\times\mathcal{R}^2 / &(l, \underline{m}, \underline{n}) \in \left( l, \frac{2 \pi m}{L_y}, \frac{2 \pi n}{L_z} \right), \text{where} \notag \\
 &(l, m, n) \in  [0, \ldots, N_x] \times [-\frac{N_y}{2},\ldots, 
 \frac{N_y}{2}-1] \times[-\frac{N_z}{2},\ldots, \frac{N_z}{2}-1] \Big\} 
 \label{eq:wavenumbermesh}
\end{align}
is the wavenumber mesh for $V_N$. The other two wavenumber meshes $\D{K}_N$ and 
$\N{K}_N$ are both equal to $K_N$ except from in index $l$ that 
ranges $[0, 1, \ldots, N_x-2]$ and $[0, 1, \ldots, N_x-4]$ for $\D{K}_N$ 
and $\N{K}_N$ respectively, see (\ref{eq:phiD}) and (\ref{eq:phiN}). We use 
notation $K_N^p(m, n)=K_N(0, m, n)$ when referring to only the two periodic 
dimensions of $K_N$.

The computational mesh in real space can be created using $N = (N_x, N_y, 
N_z)$ intervals, where the two periodic directions use uniform intervals. The 
computational mesh is given as
\begin{align}
  X_N = \Big\{ \bm{x} \in \mathcal{R}^3/&(x_i, y_j, z_k) \in \left( h(i), \frac{jL_y}{N_y}, \frac{kL_z}{N_z} \right), \text{where} \notag \\
  &(i, j, k) \in [0, 1, \ldots, N_x] \times [0, 1, \ldots, N_y-1] \times [0, 1, \ldots, N_z-1] \Big\} \label{eq:Xn}
\end{align}
where $x_i = h(i)$ represents either
\begin{equation}
 h(i) = \begin{cases}
   \cos \left(\frac{i \pi }{N_x} \right) \, &\forall \, i=0,1, \ldots, N_x \quad  \text{for Gauss-Lobatto points}, \\
   \cos \left(\frac{(2i +1)\pi}{2N_x+2} \right) \, &\forall \, i=0,1, \ldots, N_x \quad  \text{for Gauss-Chebyshev points}. \\
 \end{cases}
\end{equation}

@@

If the expansion coefficients $\hat{u}$  are known for the entire wavenumber 
mesh (\ref{eq:wavenumbermesh}), then the expression (\ref{eq:u_solx}) may be 
evaluated on the mesh (\ref{eq:Xn}) using fast Fourier and fast Shen inverse 
transforms as
\begin{align}
u(x_i, y_j, z_k, t) &= \underbrace{\frac{1}{N_z}\sum_n 
\underbrace{\frac{1}{N_y} \sum_m \underbrace{\sum_l 
\hat{u}(l,m,n)\N{\phi}_l(x_i)}_{\N{\mathcal{S}}_x^{-1}} e^{\imath \underline{m} 
y_j}}_{\mathcal{F}_y^{-1}} e^{\imath \underline{n} z_k}}_{\mathcal{F}_z^{-1}} , 
\quad \forall \, \bm{x} \in X_N, \notag \\
  u &= \N{\mathcal{T}}^{-1}(\hat{u}) =  \mathcal{F}^{-1}_z (\mathcal{F}^{-1}_y 
  (\N{\mathcal{S}}^{-1}_x (\hat{u}))), \quad \forall \, \bm{x} \in X_N.  
  \label{eq:ifft} 
\end{align}
Here $\N{\mathcal{T}}^{-1}$ is used as short notation for the complete inverse 
transform, $\mathcal{F}_{y}^{-1}$ and $\mathcal{F}_{z}^{-1}$ represent inverse 
Fourier transforms along directions $y$ and $z$ respectively, and the inverse 
Shen transform $\N{\mathcal{S}}_{x}^{-1}$ is performed along the wall normal 
$x$ direction. Note that the transforms are applied sequentially to the entire 
computational mesh. The inverse Shen transform may be computed using fast 
Chebyshev transforms for all the three bases in (\ref{eq:Tk}, \ref{eq:phiD}, 
\ref{eq:phiN}). The transforms are slightly different for the three bases and 
${\mathcal{S}}, \D{\mathcal{S}}$ and $\N{\mathcal{S}}$ are used to distinguish 
between them with obvious notation.

@@

The scalar product  $h\N{\mathcal{S}}$ in (\ref{eq:sst1}) does not represent a 
complete transform. To find a transformation between physical $u$ and spectral 
$\hat{u}$ we make use of Eq. (\ref{eq:u_solx}) directly and use the discrete 
orthogonality of the Fourier basis functions
\begin{align}
\left<u, \N{\psi}_{\bm{k}}\right>_w &= \frac{1}{N_yN_z}\left< 
\sum_{\bm{m}=(q,r,s) \in \N{K}_N} \hat{u}_{\bm{m}} \N{\psi}_{\bm{m}}, 
\N{\psi}_{\bm{k}} \right>_w, \notag \\
           &= h \sum_{q=0}^{N_x-4} \sum_{i=0}^{N_x} \N{\phi}_{q}(x_i) 
           \N{\phi}_{l}(x_i) w_i \, \hat{u}(q, {m}, {n}, t), \notag \\
           &= h \sum_{q=0}^{N_x-4} \N{B}_{lq} \hat{u}(q, {m}, {n}, t), 
           \label{eq:sst2}
\end{align}
where $\N{B}_{lq} = (\N{\phi}_q, \N{\phi}_l)_w = 
\sum_{i=0}^{N_x} \N{\phi}_{q}(x_i) \N{\phi}_{l}(x_i) w_i$ is a 
pentadiagonal mass matrix. The complete transformation is now obtained by 
setting Eq. (\ref{eq:sst2}) equal to (\ref{eq:sst1}) and solving for $\hat{u}$. 
Moving to matrix notation we get
\begin{align}
h\sum_{q=0}^{N_x-4} \N{B}_{lq}\, \hat{u}(q, {m}, {n}, t) &= 
h\N{\mathcal{S}}(u)(l, m, n, t), \notag \\
 \hat{u}_{\bm{k}}(t) &= \N{\mathcal{T}}(u) =  \N{B}^{-1}\N{\mathcal{S}}(u), 
 \quad \forall \, \bm{k} \in \N{K}_N. \label{eq:fstB}
\end{align}
where $\N{\mathcal{T}}(u)$ denotes the complete transformation, such that $u = 
\N{\mathcal{T}}^{-1}(\N{\mathcal{T}}(u))$. Note that since $\N{B}$ assembles to 
a pentadiagonal matrix the solution ($\N{B}^{-1}$) can be obtained very fast 
and 
the complete 
transformation thus requires a fast Chebyshev transform ($O(N_x \log N_x)$) and 
a fast linear algebraic solve ($O(N_x)$). Similar transforms $\mathcal{T}$ 
and $\D{\mathcal{T}}$ are defined for the two other bases (\ref{eq:Tk}) and 
(\ref{eq:phiD}), using mass matrices $B_{lq}=(T_q, T_l)_w$ (diagonal) and 
$\D{B}_{lq}=(\D{\phi}_l, \D{\phi}_q)_w$ (tridiagonal) and scalar products 
$\mathcal{S}_l(\cdot) = (\cdot, T_l)_w$ and $\D{\mathcal{S}}_l(\cdot) = (\cdot, 
\D{\phi}_l)_w$.

@@

It remains to discretize the equations in time. Kim, Moin and Mansour use 
Crank-Nicolson for the diffusive terms and Adams-Bashforth for the nonlinear 
convective terms, whereas a third order Runge-Kutta method is used for the 
nonlinear term by Moser, Kim and Mansour (1999). We have implemented both 
approaches, but for simplicity only the first one is described here. 
Discretization in time is performed using a constant time step $\triangle t$, 
such that the time is represented discretely as $t_n = n \triangle t, \, n \in 
\mathcal{Z}^{\ge}$. After 
discretizing in time, equations (\ref{eq:u1}) and (\ref{eq:g1}) are transformed 
to a matrix form that can be solved for $\hat{u}^{n+1}$ and $\hat{g}^{n+1}$ 
given wavenumbers $m$ and $n$
\begin{align}
\N{H}\hat{u}^{n+1} & = \left(2\N{A} - 2\underline{z}^2\N{B} - \N{H} 
\right)\hat{u}^{n} + \N{\mathcal{S}}(h_v^{n+1/2}) \triangle t & 
\forall \, {\bm{k}} \in \N{K}_N, 
\label{eq:ufin}\\ 
\D{H} \hat{g}^{n+1} &= \left(2 \D{B}-\D{H}\right) 
\hat{g}^{n} + \D{\mathcal{S}}(h_g^{n+1/2}) \triangle t &\forall \, 
{\bm{k}} \in 
\D{K}_N. \label{eq:gfin}
\end{align}
The coefficient matrices are given as
\begin{align}
\N{H}_{lq}(m, n) &= -\frac{\nu \triangle t}{2}\N{Q}_{lq} + \left( 1 + \nu 
\triangle t 
\underline{z}^2 \right) \N{A}_{lq} - \frac{2\underline{z}^2 + \nu \triangle t 
\underline{z}^4}{2} \N{B}_{lq} &\forall m, n \in K_N^p, \\
\D{H}_{lq}(m, n) &= -\frac{\nu \triangle 
t}{2}\D{A}_{lq} + (1 + \frac{\nu \triangle t \underline{z}^2}{2}) \D{B}_{lq} 
&\forall m, n \in K_N^p,
\end{align}
where $\underline{z}^2 = \underline{m}^2 + \underline{n}^2$. The convective 
terms are computed with a forward Adams-Bashforth projection, such that 
$\N{\mathcal{S}}(h_v^{n+1/2}) = 
\N{\mathcal{S}}(1.5h_v^{n} - 0.5 h_v^{n-1})$ and $\D{\mathcal{S}}(h_g^{n+1/2}) 
= \D{\mathcal{S}}(1.5h_g^{n} - 0.5 h_g^{n-1})$. Equations 
(\ref{eq:v0}) and (\ref{eq:w0}) are also discretized in time using 
Crank-Nicolson and Adams-Bashforth, and we obtain
\begin{align}
\left(-\frac{\nu \triangle t}{2}\D{A} + \D{B} \right)\hat{v}_0^{n+1} &= 
\left(\frac{\nu \triangle t}{2}\D{A} + \D{B} 
\right)\hat{v}^{n}_0 + \triangle t\D{B}\hat{\mathcal{H}}_{y,0}^{n+1/2} , 
\label{eq:v00} \\
\left(-\frac{\nu \triangle t}{2}\D{A} + \D{B} \right)\hat{w}_0^{n+1} &= 
\left(\frac{\nu \triangle t}{2}\D{A} + \D{B} 
\right)\hat{w}^{n}_0 + \triangle t\D{B}\hat{\mathcal{H}}_{z,0}^{n+1/2}, 
\label{eq:w00}
\end{align}
where we have used notation $\hat{v}^{n}_0 = \hat{v}(l, 0, 0, t_n)$, 
$\hat{\mathcal{H}}^{n}_{y,0} = \hat{\mathcal{H}}_y(l, 0, 0, t_n)$ and similar 
for $\hat{w}^n_0$ and $\hat{\mathcal{H}}^n_{z,0}$.

@@

The biharmonic operator $\N{H}$ is more challenging to solve efficiently, but 
it is still possible to find an algorithm that is $O(N)$ for a system of $N$ 
unknowns. Note 
that $\N{H}$ is the sum of three matrices $\N{Q}, \N{A}$ and $\N{B}$ and can be 
written as
\begin{equation}
\N{H} = c_0\N{Q} + c_1\N{A} + c_2 \N{B},
\end{equation}
where $c_0, c_1$ and $c_2$ are constants. The matrix $\N{A}_{kj}$ contains only 
three nonzero diagonals at $ j = k-2, k, k+2$,  whereas matrix $\N{B}_{kj}$ 
contains five 
nonzero diagonals at $ j = k-4, k-2, k, k+2, k+4$. The nonzero elements of the 
upper triangular $\N{Q}_{kj}$ matrix are given as
\begin{align}
 \N{Q}_{kk} &= -4(k+1)(k+2)^2 \\
 \N{Q}_{kj} &= p_kq_j + r_ks_j, \quad j = k+2, k+4, \ldots N_x-4
\end{align}
where 
\begin{align}
p_k &= 8 k (k+1)(k+2)(k+4)\pi, &q_j = \frac{1}{j+3}, \label{eq:pk} \\
r_k &= 24(k+1)(k+2)\pi, &s_j =  \frac{(j+2)^2}{(j+3)}. \label{eq:rk}
\end{align} 
For each row in $\N{H}_{kj}$ we have that $\N{H}_{kj}=c_0\N{Q}_{kj}\, \forall j 
> k+4$. A direct LU decomposition of $\N{H}$ can be performed as shown in Alg. 
\ref{alg:lu_biharmonic}, which leads to a lower triangular 
matrix $\N{L}_{kj}$ with two nonzero diagonals at $j=k-2$ and $k-4$ plus a 
unity 
main diagonal. The upper triangular matrix $\N{U}_{kj}$ contains three distinct 
diagonals at $j=k, k+2$ and $k+4$. The remaining part of $\N{U}_{kj}$ can be 
written as
\begin{equation}
\N{U}_{kj} = a_k q_j + b_k s_j, \quad j = k+6, k+8, \ldots, N_x-4 \label{eq:ab}
\end{equation}
where $\bm{a}$ and $\bm{b}$ are two new vectors that can be computed 
recursively as shown in Alg. (\ref{alg:ab}). If $\bm{a}$ and $\bm{b}$ are 
pre-computed 
the total storage requirement for the complete LU decomposition is no more than 
$7N$ and the solution of the complete system can be obtained very quickly with 
one forward elimination and a back solve. The back 
solve is very fast because of the formulation (\ref{eq:ab}), leading 
to a formula for the backwards substitution like (valid for $k \le N_x-8$)
\begin{equation}
u_k = \left(y_k - \N{U}_{k, k+2} u_{k+2} - \N{U}_{k, k+4} u_{k+4} - 
a_k\sum_{\substack{j=k+6\\k-j \text{ even}}}^{N_x-4} q_j u_j - 
b_k\sum_{\substack{j=k+6\\k-j \text{ even}}}^{N_x-4} 
s_j u_j\right)/\N{U}_{kk},
\end{equation}
that only requires 
one new addition to the row-sums for each back-traversed
row, see Alg. (\ref{alg:SolveBiharmonic}).

\begin{algorithm}
	\caption{LU decomposition of biharmonic operator $\N{H}$. Current algorithm 
	is using dense storage of $U_{kj}$, but no more than the 3 diagonals need 
	to be stored eventually.}
	\label{alg:lu_biharmonic}
	\begin{algorithmic}[1]
		\Procedure{LUbiharmonic}{}
			\State \textbf{Input:} $\{\N{H}_{kj}\}_{k, j = 0}^{N}$
	        \State $U_{0j} = \N{H}_{0j} \quad \forall \, j \in 0, 2, \ldots, N$
	        \State $U_{1j} = \N{H}_{1j} \quad \forall \, j \in 1, 3, \ldots, N$
	        \State $L_{20} = \N{H}_{20}/U_{00}$
	        \State $L_{31} = \N{H}_{31}/U_{11}$
	        \State $U_{2j} = L_{20} U_{0j} \quad \forall \, j \in 2, 4, \ldots, 
	        N$
	        \State $U_{3j} = L_{31} U_{1j} \quad \forall \, j \in 3, 5, \ldots, 
	        N$
	        \For{$k = 4, 5, \ldots, N$}
	            \State $L_{k,k-2} = {U}_{k,k-2}/{U}_{k-2, k-2}$
		        \State $U_{kj} = {U}_{kj} - L_{k, k-2} U_{k-2, j} \quad 
		        \forall j = k, k+2, \ldots, N$
		        \If{$k <= N-2$}
		        \State $L_{k+2,k-2} = \N{H}_{k+2, k-2} / U_{k-2, k-2}$
		        \State $U_{k+2,j} = \N{H}_{k+2,j} - L_{k+2, k-2} U_{k-2, j} 
		        \quad \forall j = k-2, k, \ldots, N$
		        \EndIf
	        \EndFor
	        \State \Return $\{L_{kj}\}_{k, j = 0}^{N}$, $\{U_{kj}\}_{k, j = 
	        0}^{N}$
		\EndProcedure	
	\end{algorithmic}
\end{algorithm}
